Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-19 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Revised Ch 7 and 8

In this chapter the application of two empirical algebraic flame characterizations will be compared against the results from the experiments documented in previous chapters. Two different types of applications will be used.  In the first case, relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target will be examined. The target in question is the painted gypsum wallboard.  In the second case, the height of the thermal damage on the gypsum wallboard will be examined with the use of a flame height correlation. Applications of algorithms such as these are presented in current fire dynamics and investigation textbooks as well as other application oriented documents~\cite{Icove:2013,Quintiere:1997,Asseal:2010,Beyler:1999,NUREG_1805}.  Values generated in the fire characterizations, instrumented bench scale experiments, and values from the literature will be utilized as input to predict or simulate real scale results using the algorithms.  Then the numerical results will be compared with the full-scale experimental results.  Keep in mind that with these algorithms time is not a factor.  Methods of this type are commonly referred to as ``back of the envelope'' estimates.  They are intended to provide insight as opposed to a hard and fast numerical answer.  As the experiments have shown so far there are many uncertainties which impact the results.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, ``Will the Second Item Ignite''~\cite{Babrauskas:1981}.  The idea being that if you knew the threshold heat flux that a material needed to be exposed to for the item to ignite, then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could receive enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.  This method is noted to work the best when the distance to the target is remote, at least two times the distance of the diameter of the burner away from the fire~\cite{Hamins:1991,Modak:1977}.

@@

The question is can we use this simple model to estimate the burner positions where the heat flux would be high enough to generate thermal damage on the painted gypsum board wall?  From the source fire characterizations, the average peak heat release rates from the three burners can be input into the equation along with the distance, R, and the $\chi_r$, or the radiant fraction.  R is measured from the vertical centerline of the burners so in addition to the distance between the edge of the burner and the wall, half of the burner width, 0.15~m must be added.  A typical default value used for the radiant fraction is 0.35.  Table~\ref{tab:Point_Source_HFNGG} lists the input value of the average peak heat release rate of 80~kW for both the natural gas and gasoline burner fires along with a $\chi_r$ of 0.35, and the five burner positions that were used.

The predictions provide good agreement with the natural gas results at distances greater than 1 burner width away.  Closer than one burner width array the predictions trend higher.  In the instrumented gypsum wallboard experiments the natural gas burner at $R=0.31~m$ generated a discoloration on the surface of the painted gypsum wallboard.  At $R=0.15~m$, with the burner touching the wall, enough energy was transferred to the wall fast enough that caused charring, penetration, and consumption of the painted paper surface.  Based on the data from the cone calorimeter experiments conducted as part of this study, charring, penetration, and consumption would occur somewhere between 25 and 30~kW/m$^2$.  Dillon provided a value of 26kW/m$^2$ as the minimum heat flux where ignition of the gypsum wallboard would occur~\cite{Dillon:1998}. So at $R=0.31~m$ conditions are on the edge of the threshold heat flux for ignition, both with the prediction and the measurement.  Neither value is high enough to cause the paper to char and ignite, there was only enough energy to cause discoloration.  Once the natural gas burner was moved next to the wall, the prediction and the measurement, even though they are not quantitatively in agreement, both provide a heat flux in excess of 26~kW/m$^2$.  Indicating that a fire pattern has been burned into the wall and that was in fact true.

The predictions provided in Table~\ref{tab:Point_Source_HFNGG} are significantly lower than the measured peak heat flux values from the gasoline fueled burner experiments at every burner position.  That said, the predictions indicate that at $R=0.31~m$  the heat flux is near the ignition threshold provided in the literature and at $R=0.23~m$ or less the heat flux is high enough to indicate ignition of the gypsum wallboard.  The single gasoline experiment at $R=0.31~m$ showed a peak heat flux of 57~kW/m$^2$.  The exposure was brief at this peak due to the flame movement toward the wall.  This makes the experimental values at the three closest locations to the wall difficult to compare with the point source model, and most models for that matter, since they cannot account for the flame being drawn toward the wall.  An important observation is that the fire damage on the wall was limited to an uneven horseshoe shaped pattern within 100~mm above the top of the burner edge.  This ignition was captured in Ch 4, Figure~\ref{IWGBGAS6_seq}. In other words, at $R=0.31~m$, there was an ignition of the fuel gases that were pyrolyzing from the painted gypsum board wall, but barely and in a very limited area relative to the flame shape near the surface.  At the two closer positions the char, penetration, and consumption was evident and at these locations both the predictions and the measured heat fluxes are well above the threshold ignition heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	 & Heat Flux (kW/m$^2$)     &  Nat. Gas (kW/m$^2$)      & Gasoline (kW/m$^2$) 	\\ \hline \hline
		80    &   .35 		& 0.76 	     & 3.8 $\pm$ 0.30 (8~\%) 	&  4 $\pm$ 0.32 (8~\%) 	 	   & 6 $\pm$ 0.5 (8~\%)     \\
		80    &   .35    	& 0.46	     & 10.6 $\pm$ 0.80 (8~\%) 	&  11 $\pm$ 0.88 (8~\%)  	   & 18 $\pm$ 1.4 (8~\%)    \\
		80    &   .35 		& 0.31	     & 24.0 $\pm$ 1.9 (8~\%) 	&  23 $\pm$ 1.8 (8~\%)  	   & 57 $\pm$ 4.6 (8~\%)     \\
        80    &   .35 		& 0.23	     & 43.0 $\pm$ 3.4 (8~\%) 	&  NA 					  	   & 60 $\pm$ 4.8 (8~\%)     \\
        80    &   .35 		& 0.15	     & 95.0 $\pm$ 7.6 (8~\%) 	&  55 $\pm$ 4.4 (8~\%)  	   & 61 $\pm$ 4.9 (8~\%)     \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 80~kW) with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values with heat flux data from the natural gas and the gasoline instrumented gypsum board wall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFNGG}
\end{table}

Table~\ref{tab:Point_Source_HFPUF} displays a similar development based on the average peak heat release rate, 45~kW, for the polyurethane foam burner.
Similar to the results from the gasoline comparison, the predictive heat flux values at all of the burner positions are low relative to the experimental data.  But again the results will be examined along a more qualitative basis. The experimental data provides an average peak value of 25~kW/m$^2$ for $R=0.31~m$. For the polyurethane foam burns the peak values were only averaged over a 15 second range due to the brief nature of the peak heat release rate. At $R=0.31~m$ there was no charring of the painted paper on the gypsum wallboard, only soot deposition.  No thermal damage.  At $R=0.23~m$ or less, the heat flux was high enough to indicate ignition in the experiments and that was the result.  Only the prediction at $R=0.15~m$ would definitely indicate that the gypsum wallboard would ignite based on the threshold heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	     & Heat Flux (kW/m$^2$)     &  Polyurethane (kW/m$^2$)    	\\ \hline \hline
		45       &   .35 		& 0.76 	     & 2.1 $\pm$ 0.17 (8~\%) 	&  3.5 $\pm$ 0.28 (8~\%) 	   	 \\
		45       &   .35    	& 0.46	     & 6.0 $\pm$ 0.48 (8~\%) 	&  12.3 $\pm$ 0.98 (8~\%)  	     \\
		45       &   .35 		& 0.31	     & 13.5 $\pm$ 1.1 (8~\%) 	&  25 $\pm$ 2.0 (8~\%)   	      \\
        45       &   .35 		& 0.23	     & 24.1 $\pm$ 1.9 (8~\%) 	&  36 $\pm$ 2.9 (8~\%)		      \\
        45       &   .35 		& 0.15	     & 53.5 $\pm$ 4.3 (8~\%) 	&  59 $\pm$ 4.7 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 45~kW) with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values with heat flux data from the polyurethane foam burner instrumented gypsum board wall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFPUF}
\end{table}

This was an interesting exercise but it was not very definitive.  If we were to attempt to improve the predictions by using radiative fractions that are known to be representative of the fuels, the predictions in some cases might get worse.  For example, recommended values in the literature for the $chi_r$ of natural gas range from 0.15 to 0.25 for similar sized burners~\cite{Hostikka:3,McCaffrey:1982}.  The default radiative fraction of 0.35 is in line with the values found in the literature for gasoline~\cite{Hamins:1991,McGrattan:2000}.  Literature values for polyurethane foam were elusive.

Another means to assess the point source model was to examine the heat flux data measured at 2~m away from the vertical centerline of the burners during some of the source fire characterization experiments. If we rearrange the point source equation, to solve for $\chi_r$ by using the measured average peak heat release rates and average peak heat fluxes at a distance of 2~m away the resulting value for $\chi_r$ are 0.26, 0.46 and 0.33 for the natural gas, gasoline and polyurethane burners respectively.  This distance should be more appropriate for the use of the point source model but the resultant radiative fractions for the natural gas and gasoline seem high compared to the values found in the literature.  It appears that more research is needed on radiative fractions of fire sizes and realistic fuels that might be found in a residential setting.

@@

The desire to use flame height correlations is based on the thought that the median flame height will correlate to the height of the associated fire pattern.  With the heat release rate data that we have from these experiments, comparisons will be made with the one of the more widely used empirical algorithms, known as the Heskestad flame height correlation.

@@

Heskestad, developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion.  Appropriate fuels would include liquid pool fires and other horizontal surface fires~\cite{Heskestad:SFPE,Beyler:1986}. The inputs required to use the model are the heat release rate and the diameter of the fire or the burner.

@@

Once again the average peak heat release rates from Ch 2 were be used.  The natural gas and the gasoline fueled burners both had an average peak heat release rate of 80 kW.  The polyurethane fueled burner had an average peak heat release rate of 45 kW.  Calculating the median flame heights with the average peak heat release rates measured from the three fuels yielded 1.0~m, 1.0~m, and 0.73~m for natural gas, gasoline, and polyurethane foam respectively. It should be noted that the coefficient, 0.235, is dependent on the ratio of the heat of combustion to the actual mass stoichiometric ratio of air to volatiles of the fuel.  For many liquid and gaseous fuels this coefficient is representative, Heskestad points out, however, that some common fuels, such as gasoline, deviate from the cited range. For gasoline a coefficient value of 0.2 should be used~\cite{Heskestad:SFPE}. Changing the coefficient yields a median flame height of 0.8~m for gasoline.  The assumption was made that this guidance should be followed for the polyurethane as well. Changing the coefficient for polyurethane resulted in a median flame height value of 0.57~m.

@@

\begin{table}
\centering
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
Method          &   \multicolumn{3}{|c|}{Flame Heights (m)} \\ \cline{2-4}
                &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Photographic    &       0.70 $\pm$ 0.23 (33~\%)   &	0.84 $\pm$ 0.28 (33~\%) 	& 0.46 $\pm$ 0.23 (50~\%)   \\
Video Based     &   	0.70 $\pm$ 0.11 (15~\%)   &	0.70 $\pm$ 0.1 (14~\%) 	& 0.47 $\pm$ 0.16 (34~\%)   \\
Fire Pattern    &   	0.74 $\pm$ 0.12 (16~\%)   &	0.83 $\pm$ 0.15 (18~\%) 	& 0.24 $\pm$ 0.12 (50~\%)   \\
Heskestad (range)  &   	0.94 to 1.06		    &	0.65 to 0.93        	& 0.36 to 0.72   \\
\hline
\end{tabular}
 \caption[Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions]{Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions. The measured results are shown with 95~\% confidence limits.  The Heskestad median flame height calculations are presented as a range based on the uncertainties of the average peak heat release rate of each fuel type.} 
 \label{tab:Comparing_Flame_Heights}
\end{table}

@@

The comparison of the median flame height values, the average maximum fire pattern height, and the calculated flame height range are all in agreement for the gasoline fueled pan fires. This is significant because the 95~\% confidence limits for the fire pattern and the video based flame height are both less than 20~\%.

The measured and calculated values for the polyurethane fueled burner fires are also in agreement.  The agreement in this case is a result of of the 95~\% confidence limits of the average maximum fire pattern heights being 50~\%.

Overall the Heskestad equation provided reasonable estimates of the median flame height.  The key to the successful use of any model regardless of the simplicity or complexity is using appropriate data.

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenario showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI= 5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=2.5in]{../Figures/WallTemp_10mm_0D_0_8m_guideline}
  \includegraphics[width=2.5in]{../Figures/350CThresholdwall_temp_0_8m_guideline}\\
  \hspace{4 mm}  \\ 
  \includegraphics[width=.625\columnwidth]{../Figures/GBNGreps_w_1m_guideline}\\
  \caption[FDS fire pattern surrogates based on 40~mm grid using a threshold temperature of 350~$^\circ$C compared with three replicate fire patterns from the natural gas burner positioned against the wall]{FDS fire pattern surrogates based on 40~mm grid using a threshold temperature of 350~$^\circ$C compared with three replicate fire patterns from the natural gas burner positioned against the wall.  The image with the three replicate fire patterns has a red line indicating a position approximately 1.0~m above the burner.  This height corresponds with the height of the top of the 350~$^\circ$C mark in the FDS image.}
  \label{FDS_FirePattern_comp}
\end{figure}

>>>>

In this chapter the application of two different types of algorithms was be compared against the results from the experiments documented in previous chapters.  In the first case, relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target were be examined. The target under examination was painted gypsum wallboard.  In the second case, the height of the thermal damage on the gypsum wallboard was compared with the use of a flame height correlation.

Applications of algorithms such as these are presented in current fire dynamics and investigation textbooks as well as other application oriented documents~\cite{Icove:2013,Quintiere:1997,Asseal:2010,Beyler:1999,NUREG_1805}.  Values from the source fire characterizations, instrumented bench scale experiments, and from the literature were utilized as input to predict or simulate real scale results using the algorithms.  Then the numerical results were compared with the full-scale experimental fire pattern results.  Keep in mind that with these algorithms time is not a factor.  Methods of this type are commonly referred to as ``back of the envelope'' estimates.  They are intended to provide insight as opposed to a hard and fast numerical answer.  As the experiments have shown so far there are many uncertainties which impact the results.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, ``Will the Second Item Ignite''~\cite{Babrauskas:1981}.  The data needed to use the method includes the HRR and radiant fraction from the source fire and the threshold ignition heat flux of the target material.  The idea being that if you knew the threshold heat flux that a given material needed to achieve auto ignition, then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could receive enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.  This method is noted to work the best when the distance to the target is remote, at least two times the distance of the diameter of the burner away from the fire~\cite{Hamins:1991,Modak:1977}.

@@

The question is can this simple model be used to estimate the burner positions where the heat flux would be high enough to generate thermal damage on the painted gypsum board wall?  The average peak heat release rates from the three burners, from the source fire characterizations, were input into the equation along with the distance, R, and the $\chi_r$, or the radiant fraction.  R was measured from the vertical centerline of the burners.  Therefore in addition to the distance between the edge of the burner and the wall, half of the burner width, 0.15~m was added.  As a starting value for $\chi_r$, the typical default value of 0.35 was used.  Table~\ref{tab:Point_Source_HFNGG} lists the input value of the average peak heat release rate of 80~kW for both the natural gas and gasoline burner fires along with a $\chi_r$ of 0.35, and the five burner positions that were used.

The predictions provided good agreement with the natural gas results at distances greater than 1 burner width away.  Closer than one burner width array the predictions trend higher.  In the instrumented gypsum wallboard experiments the natural gas burner at $R=0.31~m$ generated a discoloration on the surface of the painted gypsum wallboard.  At $R=0.15~m$, with the burner touching the wall, enough energy was transferred to the wall fast enough that caused charring, penetration, and consumption of the painted paper surface.  Based on the data from the cone calorimeter experiments conducted as part of this study, charring, penetration, and consumption would occur somewhere between 25 and 30~kW/m$^2$.  Dillon provided a value of 26~kW/m$^2$ as the minimum heat flux where ignition of the gypsum wallboard would occur~\cite{Dillon:1998}.  At $R=0.31~m$ the heat flux conditions were very close to the 26~kW/m$^2$ threshold heat flux for ignition, both with the prediction and the measurement.  Neither value was high enough to cause the paper to char and ignite, there was only enough energy to cause some discoloration.  Once the natural gas burner was moved next to the wall, the prediction and the measurement, even though they are not quantitatively in agreement, both provide a heat flux in excess of 26~kW/m$^2$.  Indicating that a fire pattern has been burned into the wall and that was in fact true.

The predictions provided in Table~\ref{tab:Point_Source_HFNGG} were significantly lower than the measured peak heat flux values from the gasoline fueled burner experiments at every burner position.  That said, the predictions indicate that at $R=0.31~m$  the heat flux is near the ignition threshold provided from the literature and at $R=0.23~m$ or less the heat flux is high enough to indicate ignition of the gypsum wallboard.  The single gasoline experiment at $R=0.31~m$ showed a peak heat flux of 57~kW/m$^2$.  The exposure was brief at this peak value as is was caused by movement of the flame toward the wall.  This makes the experimental values at the three closest locations to the wall difficult to compare with the point source model, and most models for that matter, since they cannot account for the flame being drawn toward the wall.  In this case, an important observation is that the fire damage on the wall was limited to an uneven horseshoe shaped pattern within 100~mm above the top of the burner edge.  This ignition was documented in Ch 4, Figure~\ref{IWGBGAS6_seq}. In other words, at $R=0.31~m$, there was an ignition of the fuel gases that were pyrolyzing from the painted gypsum board wall,which resulted in a smaller fire area of fire damage relative to the flame shaped damaged when the burner was against the wall surface.  At the two closer radii, 0.23 and 0.15~M, the char, penetration, and consumption were evident and at these locations both the predictions and the measured heat fluxes are well above the threshold ignition heat flux.  Perhaps coincidental, but at each location where the predicted heat flux was similar to the threshold from the literature, ignition of the paper on the gypsum board occurred, even though the values did not provide reasonable agreement with the from the gasoline fire measurements.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	 & Heat Flux (kW/m$^2$)     &  Nat. Gas (kW/m$^2$)      & Gasoline (kW/m$^2$) 	\\ \hline 
		80    &   .35 		& 0.76 	     & 3.9 $\pm$ 0.4 (11~\%) 	&  4 $\pm$ 0.32 (8~\%) 	 	   & 6 $\pm$ 0.5 (8~\%)     \\
		80    &   .35    	& 0.46	     & 10.5 $\pm$ 1.2 (11~\%) 	&  11 $\pm$ 0.88 (8~\%)  	   & 18 $\pm$ 1.4 (8~\%)    \\
		80    &   .35 		& 0.31	     & 23.0 $\pm$ 2.6 (11~\%) 	&  23 $\pm$ 1.8 (8~\%)  	   & 57 $\pm$ 4.6 (8~\%)     \\
        80    &   .35 		& 0.23	     & 42.0 $\pm$ 4.6 (11~\%) 	&  NA 					  	   & 60 $\pm$ 4.8 (8~\%)     \\
        80    &   .35 		& 0.15	     & 90.0 $\pm$ 10.9 (11~\%) 	&  55 $\pm$ 4.4 (8~\%)  	   & 61 $\pm$ 4.9 (8~\%)     \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 80~kW) with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values with heat flux data from the natural gas and the gasoline instrumented gypsum board wall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of HRR and heat flux values from the literature~\cite{Bryant:2003,Pitts:2006}.}
	\label{tab:Point_Source_HFNGG}
\end{table}

Table~\ref{tab:Point_Source_HFPUF} displays a similar development based on the average peak heat release rate, 45~kW, for the polyurethane foam burner.
Similar to the results from the gasoline comparison, the predictive heat flux values at all of the burner positions are low relative to the experimental data.  But again the results will be examined along a more qualitative basis. The experimental data provides an average peak value of 25~kW/m$^2$ for $R=0.31~m$. For the polyurethane foam burns the peak values were only averaged over a 15 second range due to the brief nature of the peak heat release rate. At $R=0.31~m$ there was no charring of the painted paper on the gypsum wallboard, only soot deposition.  No thermal damage.  At $R=0.23~m$ the calculated heat flux was close to but less than the threshold ignition heat flux, however in the experiments the paper ignited at this position.  Only the prediction at $R=0.15~m$ would definitely indicate that the gypsum wallboard would ignite based on the threshold heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	     & Heat Flux (kW/m$^2$)     &  Polyurethane (kW/m$^2$)    	\\ \hline 
		45       &   .35 		& 0.76 	     & 2.2 $\pm$ 0.24 (8~\%) 	&  3.5 $\pm$ 0.28 (8~\%) 	   	 \\
		45       &   .35    	& 0.46	     & 6.0 $\pm$ 0.65 (8~\%) 	&  12.3 $\pm$ 0.98 (8~\%)  	     \\
		45       &   .35 		& 0.31	     & 13.1 $\pm$ 1.4 (8~\%) 	&  25 $\pm$ 2.0 (8~\%)   	      \\
        45       &   .35 		& 0.23	     & 23.7 $\pm$ 2.6 (8~\%) 	&  36 $\pm$ 2.9 (8~\%)		      \\
        45       &   .35 		& 0.15	     & 55.7 $\pm$ 6.1 (8~\%) 	&  59 $\pm$ 4.7 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 45~kW) with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values with heat flux data from the polyurethane foam burner instrumented gypsum board wall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Bryant:2003,Pitts:2006}.}
	\label{tab:Point_Source_HFPUF}
\end{table}

This was an interesting exercise but it was not very definitive.  One means of potentially improving the predictions would be to use radiative fractions that are known to be representative of the fuels. For example, recommended values in the literature for the $chi_r$ of natural gas range from 0.15 to 0.25 for similar sized burners~\cite{Hostikka:3,McCaffrey:1982}.  The default radiative fraction of 0.35 is in line with the values found in the literature for gasoline~\cite{Hamins:1991,McGrattan:2000}.  Literature values for the radiative fraction of polyurethane foam were elusive.

Another means to assess the point source model was to examine the heat flux data measured at 2.15~m away from the vertical centerline of the burners during some of the source fire characterization experiments. By rearranging the point source equation, to solve for $\chi_r$ and by using the measured average peak heat release rates and average peak heat fluxes at a distance of 2.15~m away from the burner centerline, the resulting values for $\chi_r$ were 0.25, 0.47 and 0.33 for the natural gas, gasoline and polyurethane burners respectively.  This distance should be more appropriate for the use of the point source model but the resultant radiative fractions for the natural gas and gasoline seem high compared to the values found in the literature.

The radiant fractions determined from the measurements were applied at the two most distance locations from the wall.  At these locations it was clear that only radiative heat transfer from the fires was the only means of heat transfer.  At the closer the locations, based on the magnitude of the heat flux relative to the HRR, it appeared that convective heat transfer from the flame was playing a role.  Again this is more consistent with the use guidelines of the algorithm. The comparison between the averaged peak measured heat flux and the calculated heat flux is presented in Table~\ref{tab:Point_Source_ModChir}

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
		Fuel 		& HRR  &   $\chi_r$   & R)	     & Calculated Heat Flux      & Measured Heat Flux      \\     
					& (kW) &  			  & (m)		&  (kW/m$^2$)				&   (kW/m$^2$)                        \\ \hline 
		Nat. Gas 	& 80       &   .25 	& 0.76 	     & 2.8 $\pm$ 0.30 (11~\%) 				&  4 $\pm$ 0.32 (8~\%) 	   	 \\
		Nat. Gas 	& 80       &   .25  & 0.46	     & 7.5 $\pm$ 0.83 (11~\%) 				&  11 $\pm$ 0.88 (8~\%)  	     \\
		Gasoline 	& 80       &   .47 	& 0.76	     & 5.2 $\pm$ 0.57 (11~\%) 				&  6 $\pm$ 0.5 (8~\%)   	      \\
        Gasoline 	& 80       &   .47 	& 0.46	     & 14.1 $\pm$ 1.6 (11~\%) 				&  18 $\pm$ 1.4 (8~\%)		      \\
       Polyurethane & 45       &   .33 	& 0.76	     & 2.1 $\pm$ 0.23 (11~\%) 				&  3.5 $\pm$ 0.28 (8~\%)   	       \\
       Polyurethane & 45       &   .33 	& 0.46	     & 5.6 $\pm$ 0.62 (11~\%) 				& 12.3 $\pm$ 0.98 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values based on radiative fractions determined from the source fire characterization with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values based on radiative fractions determined from the source fire characterization with heat flux data from the instrumented gypsum board wall experiments.  Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Bryant:2003,Pitts:2006}.}
	\label{tab:Point_Source_ModChir}
\end{table}

The calculated values in Table~\ref{tab:Point_Source_ModChir} under predict the measured values.  The $\chi_r$ values determined from the source fire measurements improved the agreement between the calculated and measured heat flux values for gasoline, even though the determined values of the radiative fraction for natural gas and gasoline appear to be on the high side. However since the determined values of $\chi_r$ for natural gas and polyurethane were less that the default value of 0.35, the agreement between the calculated and measured values decreased.

The physics of the algorithm are sound, the challenge for the end user is determining the appropriate input values and making sure to use the algorithm within the appropriate distance range.  More information or research is needed on the radiative fractions of various fuels.  For example the majority of natural gas is composed of methane.  However natural gas may also contain small amounts of ethane, propane, hexane, butane, and pentane. Some of these gases have higher radiative fractions than methane. In one widely used text, a table listing radiative energy fraction values for burner diameters larger than 0.5~m, lists butane in the 20 to 40 $\%$ range and hexane in the 40 to 60 $\%$ range ~\cite{Quintiere:1997}.  So it is possible that the radiative fraction for natural gas could be higher that the radiative fraction associated with methane.  The radiative fraction value in the literature for gasoline is 0.45~\cite{Quintiere:1997, IAAI:2017}.  This is in line with the determined value.

Based on the information here, it seems that it would be difficult to use the point source method for determining the ignition of a target in a residential room and contents fire scenario.  In order to generate a heat flux greater than 20 kW/m$^2$, at a distance of 0.76~m, with a radiative fraction of 35, the HRR would have to be about 420 kW.
The fire investigation community would be well served to have additional research to examine the use of algorithms such as the point source method with realistic fuels that might be currently found in a residential setting.

@@

The desire to use flame height correlations was based on the thought that the median flame height would correlate to the height of the associated fire pattern.  With the heat release rate data that we have from these experiments, comparisons were made with the one of the more widely used empirical algorithms, known as the Heskestad flame height correlation.

@@

Heskestad, developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion.  Appropriate fuels would include liquid pool fires and other horizontal surface fires~\cite{Heskestad:SFPE,Beyler:1986}. The inputs required to use the model are the heat release rate and the diameter of the base of the fire or the burner.

@@

Once again the average peak heat release rates from Chapter 2 were be used.  The natural gas and the gasoline fueled burners both had an average peak heat release rate of 80~ kW.  The polyurethane fueled burner had an average peak heat release rate of 45~kW.  Calculating the median flame heights with the average peak heat release rates measured from the three fuels yielded 1.0~m, 1.0~m, and 0.73~m for natural gas, gasoline, and polyurethane foam respectively. It should be noted that the coefficient, 0.235, is dependent on the ratio of the heat of combustion to the actual mass stoichiometric ratio of air to volatiles of the fuel.  For many liquid and gaseous fuels this coefficient is representative, Heskestad points out, however, that some common fuels, such as gasoline, deviate from the cited range. For gasoline a coefficient value of 0.2 should be used~\cite{Heskestad:SFPE}. Changing the coefficient yields a median flame height of 0.8~m for gasoline.  The assumption was made that this guidance should be followed for the polyurethane as well. Changing the coefficient for polyurethane resulted in a median flame height value of 0.57~m.

@@

\begin{table}
\centering
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
Method          &   \multicolumn{3}{|c|}{Flame Heights (m)} \\ \cline{2-4}
                &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline 
Photographic    &       0.70 $\pm$ 0.23 (33~\%)   &	0.84 $\pm$ 0.28 (33~\%) 	& 0.46 $\pm$ 0.23 (50~\%)   \\
Video Based     &   	0.70 $\pm$ 0.11 (15~\%)   &	0.70 $\pm$ 0.1 (14~\%) 	& 0.47 $\pm$ 0.16 (34~\%)   \\
Fire Pattern    &   	0.74 $\pm$ 0.12 (16~\%)   &	0.83 $\pm$ 0.15 (18~\%) 	& 0.24 $\pm$ 0.12 (50~\%)   \\
Heskestad (range)  &   	0.94 to 1.06		    &	0.65 to 0.93        	& 0.36 to 0.72   \\
\hline
\end{tabular}
 \caption[Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions]{Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions. The measured results are shown with 95~\% confidence limits.  The Heskestad median flame height calculations are presented as a range based on the uncertainties of the average peak heat release rate of each fuel type.} 
 \label{tab:Comparing_Flame_Heights}
\end{table}

@@

For the gasoline fueled fires, the comparison of the median flame height, the average maximum fire pattern height, and the calculated flame height range values are all in agreement. This is significant because the 95~\% confidence limits for the fire pattern and the video based flame height are both less than 20~\%.

The measured and calculated values for the polyurethane fueled burner fires are also in agreement.  However the agreement in this case is largely the result of the broad 95~\% confidence limits of the range of the average peak heat release rate of the polyurethane foam.  The burning variation of polyurethane foam resulting in the wide range of flame heights and patterns with 95~\% confidence limits of $\pm$ 50~\%.

Overall the Heskestad equation provided reasonable estimates of the median flame height.  The key to the successful use of any model regardless of the simplicity or complexity is using the appropriate data and understanding the basis of the algorithm, in order to assess any underlying assumptions.

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenario showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI=5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The program Smokeview uses the numerical results of FDS to create scientific visualizations.  In this section, comparisons were made between Smokeview generated images and photographs of fire patterns. The criteria for the `charred' area of the gypsum wall board was a threshold temperature of 360~$^\circ$C.  This value was chosen based on the ignition temperature data collected as part of this study and documented in Chapter~5.

Keep in mind that this ignition temperature is lower, by more than 150~$^\circ$C, than other values for the ignition temperature of gypsum wallboard in the literature~\cite{Dillon:1998,Babrauskas:2003}.  It is possible that the paint on the surface of the samples used in this study, had an impact or that during the decades between the studies the properties of the material have been changed in someway.  Thermal properties data on the gypsum board paper used in the study was not available from the manufacturer.

The comparisons were made for the natural gas burner fire positioned against a painted gypsum board wall.  Figure~\ref{FDS_FirePattern_comp} displays two Smokeview images.  In the image on the left the isothermal areas of the wall shown with a temperature range of 20~$^\circ$C to greater than 520~$^\circ$C.  The Smokeview image on the right has the `char' feature applied with the threshold temperature of 360~$^\circ$C to coincide with cone calorimeter ignition temperature data. The upper edge of the char area is approximately 1~m above the burner.  The upper edge of the 520~$^\circ$C area in the left image was approximately 0.8~m above the burner. This images were generated with the RI=10 model which has grid cells that were 0.4~m in size. The FDS images are compared with three photographs of fire patterns from the compartment fire pattern experiments documented in Chapter~6.  The three images from left to right exhibit a range of fire pattern heights from approximately 0.7 to 0.8~m.  The average height of the natural gas fire patterns from this open wall construction series was 0.74 with a 95~\% certainty range of $pm$~0.12~m. The simulated fire pattern is taller than the actual patterns.  However is the upper ignition temperature was used the height values between the actual and the simulated would have been similar.  The RI=10 simulation was used as a means to optimize the graphical resolution, even though the temperature range of the flames were shifted up relative to the RI=5 case.
